---
title: "ESM 244 Lab 1"
author: "I. Requena"
date: "2023-01-13"
output: html_document
---

```{r setup, include=TRUE, echo= TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(here)
library(sf)
library(tmap)

```

### Read in the "trees data"
```{r}
sf_trees <- read_csv(here('data','sf_trees','sf_trees.csv'))

# summary(sf_trees)
# names(sf_trees) 

# names gives us the columns names

sf_trees_status <- sf_trees %>% 
  group_by(legal_status) %>% 
  summarise(tree_count = n())

top_5_status <- sf_trees %>% 
  group_by(legal_status) %>% 
  summarise(tree_count = n()) %>% 
  slice_max(tree_count, n= 5) %>% 
  arrange(-tree_count)

# slice max will find the maximum of a variable and keep some of the values (in this case it keeps the top 5)

# arrange will do from low to high by defect. If you want to invert it, put a minus - sign before tree_count arrange(-tree_count)
```

### Plot "Top 5 Legal Status"
```{r}
ggplot(data = top_5_status, aes(x = fct_reorder (legal_status, tree_count), y = tree_count))+
  geom_col() +
  labs(y="Tree count", x="Legal status") +
  coord_flip()+
  theme_minimal()

# geom_col() makes a bar chart, where its height is as high as the highest y bar.

#these are not meaningful, so lets rearrange it with fct_reorder. Before it was aes(x=legal_status, y= tree_count)

# labs() is labels
# coord_flip() flip x and y axis

```


### Keep observations where legal status is "Permitted Site" and caretaker is "MTA"

```{r}
# sf_trees$legal_status %>% unique()
permitted_mta <- sf_trees %>% 
  filter(legal_status == "Permitted Site" & caretaker == "MTA")

# sf_trees$legal_status %>% unique() is to view the variables

# == means it matches " "
```

### Keep Blackwood acacia trees, only legal_status, date, latitude, longitude columns

```{r}
blackwood_acacia <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% 
  select(legal_status, date, latitude, longitude) 
         
# str_detects is string dectects where it looks for a pattern

# make a plot
ggplot(data=blackwood_acacia, aes(x=longitude, y=latitude))+
  geom_point()
```

### Use tidyr::separate() to separate one column into multiple. Think species column of original data. Use tidyr::unite() to rejoin
```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c('spp_scientific', 'spp_common'), sep = ' :: ')

# we can reverse this, where we can join to columns (think joining genus and species)

sf_trees_unite <- sf_trees %>% 
  unite("id_status", tree_id:species, sep = '_NEW_')
```

### Make some actual maps!

```{r}
blackwood_acacia_sf <- blackwood_acacia %>% 
  drop_na(longitude,latitude) %>% 
  st_as_sf(coords = c('longitude', 'latitude'))


  
# sf=spatial feature
# st_as_sf() is to take spatial and temporal data and turn it into an sf geometry feature 

st_crs(blackwood_acacia_sf) <- 4326
# st_crs()=is cordinate reference system
# 4326 numeric code for a reference system

ggplot(data=blackwood_acacia_sf)+
  geom_sf(color='darkgreen')+
  theme_minimal()

# geom_sf()= plots simple features. It looks for geometry column automatically
```

#### read in SF Streets data
```{r}
sf_map_sf <- read_sf(here('data', 'sf_map', 'tl_2017_06075_roads.shp')) %>% 
  st_transform(4326)
  
# sf_map_sf= san francisco map in simple features
 # st_transform()= transforms coord systems

#st_crs(sf_map_sf)

# st_crs= tells us some info on reference system
```

### Map this out!
```{r}
ggplot()+
  geom_sf(data = sf_map_sf, size = 0.1, color = 'darkgrey')+
  geom_sf(data = blackwood_acacia_sf, color = 'red', size = 0.5)+
  theme_void()+
  labs(title = 'Blackwood acacias in San Francisco')

# theme_void()=cleans out the map
```

### Interactive 
```{r}
tmap_mode('view')
tm_shape(blackwood_acacia_sf) + 
  tm_dots()
```

























